Competing Constructions for German Measure Phrases

### 2017-03-15 Latex/article.tex
some work on previous analyses

\usepackage[small]{dgruyter}
\usepackage{microtype}

@@

I use the term \textit{measure noun phrase} (MNP) to refer to noun phrases (NPs) where a kind-denoting (count or mass) noun depends on another noun specifying a quantity of the objects or the substance denoted by the kind noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.%
\footnote{It would be more technically appropriate -- but also more cumbersome -- to speak of the \textit{kind noun phrase} because the embedded noun can be modified, cf.\ below.}
Measure nouns can be all sorts of nouns which genuinely denote a quantity (such as \textit{litre} or \textit{amount}) but also nouns denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \cite[284]{Brems2003} also considers nouns as measure nouns \textit{which, strictly speaking, do not designate a `measure', but display a more nebulous (sic!) potential for quantification}.

Descriptively, three different situations need to be distinguished \wrt case assignment in German measure noun phrases.
If the kind noun comes with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{Wheras partitives are constructions denoting a proper part-of relation such as \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities such as \textit{a sip of wine}.
See the references just given.}
I use * to denote general unacceptability, and the acceptability judgements given in (\ref{ex:intro:pseudopartitive}) are completely undisputed.
Notice that the kind noun is obligatory in the genitive and that case agreement with the measure noun is not an option (but see immediately below).
I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

\begin{enumerate}
  \item{\label{it:intro:datsg} alternative forms of dative singular neuter adjectives}
  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} the parallel plural/collective construction(s)}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

@@

Turning to \ref{it:intro:plurl}, readers might habe notices that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun with a collective denotation such as \textit{ein Sack kleine Äpfel\slash kleiner Äpfel} `a bag of small apples' (see Table~\ref{tab:syncretisms} for the inflection), a similar alternation between PGC and NAC can be observed.%
\footnote{On the collective interpretation of such constructions, see \cite{Eschenbach1994}.}
In line with experimental results reported in \cite[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (over 92\%, cf.\ Section~\ref{sec:corpusmodels}) that the alternation is virtually unanalysable.
While this will play a role in the interpretation of the corpus findings, such plural\slash collective MNPs will not be dealt with prominently in the remainder of the paper.

Turning to \ref{it:intro:noifl}, some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent'.
I treat them like any other measure noun because they enter into both the NAC as in (\ref{ex:uninflectedmeasures:a}) and the PGC as in (\ref{ex:uninflectedmeasures:b}).
The statistical models presented in Section~\ref{sec:corpusmodels} will, however, show whether they (or other highly grammaticalised measure nouns) favor the NAC or the PGC.

@@

As stated in Section~\ref{sec:cogocl}, I consider it essential not to disregard generalisations discovered in descriptive, generative or formal frameworks, even if the accompanying formulations do not fit into probabilistic settings.
This section therefore reviews some existing analyses.
First, a major interest in generatively oriented linguistics (and formal semantics) is to determine the syntactic structure of the constructions in question.
In the case at hand, the question

% left/right-headed (Vos: DPC and DCC)
% role of the adjective in the NP
% everything-as-measure-noun and grammaticalisation (Brems)
% factors influencing (Hentschel)
% some remarks on the history: Zimmer

% Cite
% q-as-a-functional-category-loebel
% eschenbach__massangaben-im-kontext_KOMMENTIERT
% hentschel_1993_zfgl.1993.21.3.320
% loebel__apposition-in-der-quantifizierung
% Bergenholtz_1985_PPN345203720_0107_log9
% brems_ijcl2003-measure-nouns
% [] FUNKTIONALE KATEGORIEN IN NOMINALPHRASEN
% gallmann-2004
% Rutkowski_Grammaticalized_Partitives_Pseudo-Partitives
% Vater_H_2006

% Brems (Language Sciences 53 (2016) 160–176) also on grammaticalisation of SINGLE measure noun.

%\footnote{More work on the syntax of such constructions from various theoretical points of view can be found in \cite{Bhatt1990}, \cite{GallmannLindauer1994}, \cite{Demske2001}, \cite{Gerstenberger2015}, etc.}
%\footnote{A similar broad notion of measure noun is used, for example, in \cite{Brems2003}.}
%\footnote{In many cases, a construction with \textit{von} `of' is also possible and maybe even more natural.
%I do not consider this construction because it is structurally different from the cases where the alternation occurs, cf.\ below.
%Interestingly, constructions of the type \textit{NP of\slash von NP} are the ones mostly discussed in the literature on (pseudo-)partitives and }

@@

\begin{enumerate}
  \item\label{enum:data:step1} bootstrapping a list of the one hundred most frequent mass nouns,
  \item\label{enum:data:step2} bootstrapping a list of all measure nouns with which the mass nouns co-occur in the unambiguous measure construction with a determiner, and
  \item\label{enum:data:step3} sampling the target constructions by querying each combination of mass noun and measure noun found in step \ref{enum:data:step2}.
\end{enumerate}

@@

\subsection{Descriptive statstics and non-hierarchical models}
\label{ssec:corpusnonhierarchicalmodel}

@@

% Why didn't we use "mit einem Liter guten Benzin"?
% Generally noisy sample, impossible to tell whether it's an omitted genitive -s on the noun!
% If ChZ did automatic queries, most likely incorrect results because of noise.

@@

\pagebreak

>>>>

\usepackage[small]{dgruyter}
\usepackage{microtype}
\usepackage{enumerate}

@@

I use the term \textit{measure noun phrase} (MNP) to refer to noun phrases (NPs) where a kind-denoting (count or mass) noun depends on another noun specifying a quantity of the objects or the substance denoted by the kind noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.%
\footnote{It would be more technically appropriate -- but also more cumbersome -- to speak of the \textit{kind noun phrase} because the embedded noun can be modified, cf.\ below.}
Measure nouns can be all sorts of nouns which genuinely denote a quantity (such as \textit{litre} or \textit{amount}) but also nouns denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \cite[284]{Brems2003} also considers nouns as measure nouns \textit{which, strictly speaking, do not designate a `measure', but display a more nebulous} (\textit{sic!}) \textit{potential for quantification}.

Descriptively, three different situations need to be distinguished \wrt case assignment in German measure noun phrases.
If the kind noun comes with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{Whereas partitives are constructions denoting a proper part-of relation such as \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities such as \textit{a sip of wine}.
See the references just given.
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \citep{Zimmer2015} while some realise the difference \citep[225]{Eschenbach1994}.}
I use * to denote general unacceptability, and the acceptability judgements given in (\ref{ex:intro:pseudopartitive}) are completely undisputed.
Notice that the kind noun is obligatory in the genitive and that case agreement with the measure noun is not an option (but see immediately below).
I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

\begin{enumerate}[i.]
  \item{\label{it:intro:datsg} alternative forms of dative singular neuter adjectives}
  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} the parallel plural/collective construction(s)}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

@@

Turning to \ref{it:intro:plurl}, readers might habe notices that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun with a collective denotation such as \textit{ein Sack kleine Äpfel\slash kleiner Äpfel} `a bag of small apples' (see Table~\ref{tab:syncretisms} for the inflection), a similar alternation between PGC and NAC can be observed.%
\footnote{On the collective interpretation of such constructions, see \cite{Eschenbach1994}.
It is, however, very surprising that she does not notice the morpho-syntactic distribution of PGC and NAC, even when she gets very close to listing the relevant data on p.\ 217.}
In line with experimental results reported in \cite[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (over 92\%, cf.\ Section~\ref{sec:corpusmodels}) that the alternation is virtually unanalysable.
While this will play a role in the interpretation of the corpus findings, such plural\slash collective MNPs will not be dealt with prominently in the remainder of the paper.

Turning to \ref{it:intro:noifl}, some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent'.%
\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass'.
A non-inflected variant \textit{zwei Glas roter Wein} and an inflected variant \textit{zwei Gläser roter Wein} `two glasses of wine' co-exist.
The non-inflected variant denotes the quantity of wine fitting into two glasses, the inflected variant denotes two glasses filled with wine.}
I treat these cases like any other measure noun because they enter into both the NAC as in (\ref{ex:uninflectedmeasures:a}) and the PGC as in (\ref{ex:uninflectedmeasures:b}).
The statistical models presented in Section~\ref{sec:corpusmodels} will, however, show whether they (or other highly grammaticalised measure nouns) favor the NAC or the PGC.

@@

As stated in Section~\ref{sec:cogocl}, I consider it essential not to disregard generalisations discovered in descriptive, generative or formal frameworks, even if the accompanying formulations do not fit into probabilistic settings.
This section therefore reviews some existing analyses of the OGC-NAC alternation or related issues, including generative approaches.

% Grammaticalisation (Brems, Rutkowski)
% Syntactic structure: headedness (Galmann>Löbel, Eschenbach p. 213, Vos: DPC/DCC), difference to English (Brems)
% Syntactic structure: adjectives-as-determiners and layers in NP, role assignment to nouns in PGC/NAC
%     Gallmann>Bhatt1990 DP[gen] vs NP[agree]
% Influencing factors: Hentschel, Zimmer
% Squeeze in Anttila & Fong

A major interest in generatively oriented linguistics (and formal semantics) is to determine the syntactic structure of the constructions in question.
In the case at hand, the question

@@

\begin{enumerate}[i.]
  \item\label{enum:data:step1} bootstrapping a list of the one hundred most frequent mass nouns,
  \item\label{enum:data:step2} bootstrapping a list of all measure nouns with which the mass nouns co-occur in the unambiguous measure construction with a determiner, and
  \item\label{enum:data:step3} sampling the target constructions by querying each combination of mass noun and measure noun found in step \ref{enum:data:step2}.
\end{enumerate}

@@

\subsection{Descriptive statistics and non-hierarchical models}
\label{ssec:corpusnonhierarchicalmodel}

@@

@@
